Annotated example of index standardization with VAST

### 2022-04-07 doc/appendix-pcod-VAST.Rmd
Appendix edits
Update VAST comparison; delta

---
title: "Annotated example of index standardization with VAST"
output:
  pdf_document:
    highlight: tango
---

@@

settings <- make_settings(
  n_x = 205, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = TRUE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

sdmTMB does not have built-in delta models; however, we can fit the two components separately and combine the predictions. Here, we will fit the binomial presence-absence model and then a model for the positive catches only with a Gamma observation distribution and a log link. We will use the VAST-created mesh so the results are comparable.

```{r sdmTMB-example, echo = TRUE, cache=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
  mesh = fit$spatial_list$MeshList$isotropic_mesh)
tictoc::tic()
m1 <- sdmTMB(present ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh, family = binomial(link = "logit"),
  time = "year", spatiotemporal = "iid", spatial = "on")
tictoc::toc()
m1

dat2 <- subset(pcod, density > 0)
mesh2 <- make_mesh(dat2, xy_cols = c("X", "Y"),
  mesh = fit$spatial_list$MeshList$isotropic_mesh)
tictoc::tic()
m2 <- sdmTMB(density ~ 0 + as.factor(year), 
  data = dat2, mesh = mesh2, family = Gamma(link = "log"),
  time = "year", spatiotemporal = "iid", spatial = "on")
tictoc::toc()
m2
```

```{r compare-vast-sdmTMB1, echo = FALSE}
plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft") {
  s <- vast_model$parameter_estimates$SD
  vast_est1 <- as.list(s, "Estimate", report = FALSE)
  vast_est2 <- as.list(s, "Estimate", report = TRUE)
  vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
  vast_sd2 <- as.list(s, "Std. Error", report = TRUE)
  sdmTMB_est <- as.list(sdmTMB_model$sd_report, "Estimate", report = FALSE)
  sdmTMB_sd <- as.list(sdmTMB_model$sd_report, "Std. Error", report = FALSE)
  b_year_vast <- vast_est1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  b_year_vast_se <- vast_sd1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  years <- sort(unique(pcod$year))
  lwr_vast <- b_year_vast - 2 * b_year_vast_se
  upr_vast <- b_year_vast + 2 * b_year_vast_se
  plot(years, b_year_vast, ylim = range(c(lwr_vast, upr_vast)))
  segments(years, lwr_vast, years, upr_vast)
  years <- years + 0.05
  points(years, sdmTMB_est$b_j)
  segments(years, sdmTMB_est$b_j - 2 * sdmTMB_sd$b_j, 
    years, sdmTMB_est$b_j + 2 * sdmTMB_sd$b_j,
    col = "red")
  legend("topright", legend = c("VAST", "sdmTMB"), 
    col = c("black", "red"), bty = "n", lty = c(1, 1))
}
```

@@

```{r compare-vast-sdmTMB2, echo = TRUE}
plot_betas(fit, m1, "beta1_ft")
plot_betas(fit, m2, "beta2_ft")
```

@@

```{r sdmTMB-sim-index, cache=TRUE, echo=TRUE}
tictoc::tic()
set.seed(28239)
p_bin <- predict(m1, newdata = qcs_grid, sims = 500)
p_pos <- predict(m2, newdata = qcs_grid, sims = 500)

p_bin_prob <- plogis(p_bin)
p_pos_exp <- exp(p_pos)
p_combined <- log(p_bin_prob * p_pos_exp)
ind_combined_sims <- get_index_sims(p_combined, area = rep(4, nrow(p_combined)))
tictoc::toc()
```

@@

```{r index-compare, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
vast_i1 <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
    se = Std..Error.for.ln.Estimate.) %>% 
  select(index, year, est, se) %>% 
  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
  mutate(upr = exp(log(est) + qnorm(0.975) * se))
sdm_i <- ind_combined_sims %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(alpha = 0.3) +
  geom_line() + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE)
```

```{r}
plot(ind_sdmTMB$est, vast_i$est);abline(0, 1)
plot(ind_sdmTMB$upr, vast_i$upr);abline(0, 1)
plot(ind_sdmTMB$lwr, vast_i$lwr);abline(0, 1)

(ind_sdmTMB$est - vast_i$est) / vast_i$est
(ind_sdmTMB$upr - vast_i$upr) / vast_i$upr
(ind_sdmTMB$lwr - vast_i$lwr) / vast_i$lwr
```

>>>>

---
title: "Annotated example of index standardization with VAST"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
---

@@

settings <- make_settings(
  n_x = 205, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = FALSE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

TODO sdmTMB:

```{r sdmTMB-example, echo = TRUE, cache=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
  mesh = fit$spatial_list$MeshList$isotropic_mesh)

tictoc::tic()
fit_sdmTMB <- sdmTMB(
  density ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh, 
  family = delta_gamma(),
  time = "year", silent = FALSE,
  control = sdmTMBcontrol(newton_loops = 1)
)
tictoc::toc()
fit_sdmTMB
```

```{r compare-vast-sdmTMB1, echo = FALSE}
plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft", sdmTMB_pars = 1) {
  s <- vast_model$parameter_estimates$SD
  vast_est1 <- as.list(s, "Estimate", report = FALSE)
  vast_est2 <- as.list(s, "Estimate", report = TRUE)
  vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
  vast_sd2 <- as.list(s, "Std. Error", report = TRUE)
  sdmTMB_est <- as.list(sdmTMB_model$sd_report, "Estimate", report = FALSE)
  sdmTMB_sd <- as.list(sdmTMB_model$sd_report, "Std. Error", report = FALSE)
  b_year_vast <- vast_est1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  b_year_vast_se <- vast_sd1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  years <- sort(unique(pcod$year))
  lwr_vast <- b_year_vast - 2 * b_year_vast_se
  upr_vast <- b_year_vast + 2 * b_year_vast_se
  plot(years, b_year_vast, ylim = range(c(lwr_vast, upr_vast)))
  segments(years, lwr_vast, years, upr_vast)
  years <- years + 0.05
  points(years, sdmTMB_est$b_j[,sdmTMB_pars])
  segments(years, sdmTMB_est$b_j[,sdmTMB_pars] - 2 * sdmTMB_sd$b_j[,sdmTMB_pars], 
    years, sdmTMB_est$b_j[,sdmTMB_pars] + 2 * sdmTMB_sd$b_j[,sdmTMB_pars],
    col = "red")
  legend("topright", legend = c("VAST", "sdmTMB"), 
    col = c("black", "red"), bty = "n", lty = c(1, 1))
}
```

@@

```{r compare-vast-sdmTMB2, echo = TRUE}
par(mfrow = c(2, 1), cex = 0.8, mar = c(1.5, 1, 1, 1), oma = c(2, 3, 1, 1))
plot_betas(fit, fit_sdmTMB, "beta1_ft", sdmTMB_pars = 1)
plot_betas(fit, fit_sdmTMB, "beta2_ft", sdmTMB_pars = 2)
```

@@

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
tictoc::tic()
p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE,
  area = rep(4, nrow(qcs_grid))
)
ind <- get_index(p, bias_correct = FALSE)
tictoc::toc()
```

@@

```{r index-compare, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
    se = Std..Error.for.ln.Estimate.) %>% 
  select(index, year, est, se) %>% 
  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
  mutate(upr = exp(log(est) + qnorm(0.975) * se))
sdm_i <- ind %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(alpha = 0.1) +
  geom_line(alpha = 0.8) + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE)
```

```{r}
plot(ind$est, vast_i$est[vast_i$est != 0]);abline(0, 1)
plot(ind$upr, vast_i$upr[vast_i$est != 0]);abline(0, 1)
plot(ind$lwr, vast_i$lwr[vast_i$est != 0]);abline(0, 1)

(ind$est - vast_i$est[vast_i$est != 0]) / vast_i$est[vast_i$est != 0]
(ind$upr - vast_i$upr[vast_i$est != 0]) / vast_i$upr[vast_i$est != 0]
(ind$lwr - vast_i$lwr[vast_i$est != 0]) / vast_i$lwr[vast_i$est != 0]
```
